her---
title: "Lab02-Acosta_yc edits"
output: html_document
author: yc edits
date: 03-30-22
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(stringr)
```

```{r}
# store data dictionary file path
DD_FILEPATH <- here::here( "/Users/jestrii98/Desktop/CPP 528/Week 2/data/rodeo/ltdb_data_dictionary.csv" )

# import data dictionary
dd <- read.csv( DD_FILEPATH, stringsAsFactors=F )
```

# ONE: Filter variables by theme or group.

```{r}
age <- c (str_detect(dd$definition, regex("years old", ignore_case = T)),
           str_detect(dd$definition, regex("age", ignore_case = T)),
           str_detect(dd$definition, regex("persons age", ignore_case = T)))
           

age <- age %>%
  cbind.data.frame(dd, age) %>%
  filter(age == T) 

View(age)
```

# another example

```{r}
dd %>% filter(str_detect(definition, "income")) %>%
  slice()


# The function should return all of the rows of the dataframe that belong to the group.

age.search <- function(age.string) {
  vector <- age
  these <- grepl (age.string, vector, ignore.case = T)
  dat.sub.age <- age [these ]
  return( dat.sub.age)
}

age.search("age")
```

# TWO: create a function that searches variable descriptions for a specific string and returns any

# that match.

```{r}
varsearch <- function( string)
{
  vector <- dd$definition 
  these <- grepl( string, vector, ignore.case=T )
  dat.sub <- dd[ these, ]
  return( dat.sub )
}
varsearch("income")
```

# Create a function to filter variables by time periods.

#Specifically, the user will specify the time periods of interest for the study and the #function will identify all variables that have measures for those periods.

```{r}
# #variable formation:
seventies <- c (grepl("[1-9]",dd$X1970.f),
              grepl("[1-9]",dd$X1970.s))

variables.1970s <- cbind (dd,seventies) %>%
filter(seventies == T)
# 
# 
# 
eighties <- c(grepl("[1-9]",dd$X1980.f),
               grepl("[1-9]",dd$X1980.s))
# 
variables.1980s <- cbind (dd,eighties) %>%
filter(eighties == T)
# 
# 
# 
# 
nineties <- c (grepl("[1-9]",dd$X1990.f),
 grepl("[1-9]",dd$X1990.s))
# 
variables.1990s <- cbind (dd,nineties) %>%
 filter(nineties == T)
# 
# 
# 
# 
early.2000s <- c(grepl("[1-9]",dd$X2000.f),
                  grepl("[1-9]",dd$X2000.s))
# 
 variables.2000s <- cbind (dd,early.2000s) %>%
   filter(early.2000s == T)
# 
# 
# 
# 
 later.2000s<- c(grepl("[1-9]",dd$X2010.f),
                 grepl("[1-9]",dd$X2010.s))
# 
 variables.2010s <- cbind (dd,later.2000s) %>%
  filter(later.2000s == T)
# # choose from time periods:
 variables.1970s
 variables.1980s
 variables.1990s
 variables.2000s
 variables.2010s
# 
# # function
 find.variables.for.time.period <- function(time.period) {
   variables <- time.period %>% 
     select(4)
   return(variables)
 }
# 
# 
# 
 find.variables.for.time.period(variables.2010s)
```

#need to remove NAs and blanks from function

#BONUS: create a function that adds a column to the current LTDB dataset \# does not work don't fool yourselves. just playing around at 3am
