her---
title: "Lab02-Acosta_yc edits"
output: html_document
author: yc edits
date: 03-30-22
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(stringr)
library(tidyr)
library(pander)
```

```{r}
# store data dictionary file path
DD_FILEPATH <- here::here( "data/rodeo/ltdb_data_dictionary.csv" )

# import data dictionary
dd <- read.csv( DD_FILEPATH, stringsAsFactors=F )
source("/Users/jestrii98/RStudio_All/CPP-528-Project/labs/wk02/Utilities.R")
```

# ONE: Filter variables by theme or group.

```{r}
vector <- dd$category 
string <- "age"

these <- grepl( string, vector, ignore.case=T )
dd.sub <- dd[ these, ]
dd.sub  

catsearch <- function( string, dd )
{
  vector <- dd$category
  these <- grepl( string, vector, ignore.case=T )
  dd.sub <- dd[ these, ]
  return( dd.sub )
}

```
```{r}
catsearch( string="age", dd )
```

# another example

```{r,echo=FALSE,results = FALSE}
vector <- dd$definition 
string <- "income"


these <- grepl( string, vector, ignore.case=T )
dd.sub <- dd[ these, ]
dd.sub  # print it

varsearch <- function( string, dd )
{
  vector <- dd$definition
  these <- grepl( string, vector, ignore.case=T )
  dd.sub <- dd[ these, ]
  return( dd.sub )
}

```

```{r}
varsearch( string="income", dd)
```

# TWO: create a function that searches variable descriptions for a specific string and returns any

# that match.

```{r}
varsearch <- function( string)
{
  vector <- dd$definition 
  these <- grepl( string, vector, ignore.case=T )
  dat.sub <- dd[ these, ]
  return( dat.sub )
}
varsearch("income")
```

# Create a function to filter variables by time periods.

#Specifically, the user will specify the time periods of interest for the study and the #function will identify all variables that have measures for those periods.

```{r,echo=FALSE,results = FALSE}
#variable formation:
seventies <- c (grepl("[1-9]",dd$X1970.f),
                grepl("[1-9]",dd$X1970.s))

variables.1970s <- cbind (dd,seventies) %>%
  filter(seventies == T)



eighties <- c (grepl("[1-9]",dd$X1980.f),
               grepl("[1-9]",dd$X1980.s))

variables.1980s <- cbind (dd,eighties) %>%
  filter(eighties == T)




nineties <- c (grepl("[1-9]",dd$X1990.f),
               grepl("[1-9]",dd$X1990.s))

variables.1990s <- cbind (dd,nineties) %>%
  filter(nineties == T)




early.2000s <- c(grepl("[1-9]",dd$X2000.f),
                 grepl("[1-9]",dd$X2000.s))

variables.2000s <- cbind (dd,early.2000s) %>%
  filter(early.2000s == T)




later.2000s<- c(grepl("[1-9]",dd$X2010.f),
                grepl("[1-9]",dd$X2010.s))

variables.2010s <- cbind (dd,later.2000s) %>%
  filter(later.2000s == T)
# choose from time periods:
# variables.1970s
# variables.1980s
# variables.1990s
# variables.2000s
# variables.2010s

# function
find.variables.for.time.period <- function(time.period) {
  variables <- time.period %>% 
    select(4)
  return(variables)
}

```
```{r}
# choose from time periods:
# variables.1970s
# variables.1980s
# variables.1990s
# variables.2000s
# variables.2010s
(find.variables.for.time.period(variables.1990s))
```




#BONUS: create a function that adds a column to the current LTDB dataset \# does not work don't fool yourselves. just playing around at 3am
