---
title: "Lab 05"
author: "TEAM 1"
date: '2022-04-12'
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, fig.align='center', error=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

```{r, echo=FALSE, message=FALSE}
options(warn=-1)
# load necessary packages ----
library( dplyr )
library( knitr )
library( pander )
library( stargazer )
library( scales )
library(tidyverse)
library(data.table)
```


```{r message=FALSE, echo=FALSE}
# load necessary functions and objects ----

# use import::source line below when not knitting

# import::source(here::here("labs/wk05/lab_05_source.R"))

## note: because of issues with masking here functions and import being a package that does not dance well with knitting, having the libraries load in this order and in this chunk has seemed to help with knitting and loading errors. 

# note: all of these are R objects that will be used throughout this .rmd file
library(import)
library(here)
import:::here ("S_TYPE",
            "INFLATION_RATE",
             "d",
             "df",
             "d3",
             "PLOTS",
             "%>%",
             # notice the use of here::here() that points to the .R file
             # where all these R objects are created
          .from = here::here("labs/wk05/lab_05_source.R"),
          .character_only = TRUE)

detach("package:import", unload = TRUE)
```

### Full dataset: first 15 columns


```{r message=FALSE, results='asis', echo=FALSE}

pander(data.table(d[1,1:15]))


```


### Home Values - Summary Statistics



```{r, results='asis', echo=FALSE}
# show summary statistics
stargazer::stargazer( df, 
                      type=S_TYPE,
                      digits=0, 
                      summary.stat = c("min", "p25","median","mean","p75","max") )
```





```{r message=FALSE, echo=FALSE}
### POVERTY RATES
gridExtra::grid.arrange( PLOTS$pov_rate_2000$nmtc, 
                         PLOTS$pov_rate_2000$lihtc, 
                         nrow = 1 )
```

```{r message=FALSE, echo=FALSE}
### HOME VALUES


gridExtra::grid.arrange( PLOTS$mhv_2000$nmtc, 
                         PLOTS$mhv_2000$lihtc, 
                         nrow = 1 )
```

```{r message=FALSE, echo=FALSE}
### MHV Growth Rates (DV in Model)
gridExtra::grid.arrange( PLOTS$mhv_growth$lihtc, 
                         PLOTS$mhv_growth$nmtc, 
                         nrow = 1 )
```



```{r message=FALSE, echo=FALSE}
### Variables and Wrangling

# remember to use here::here() when importing data
d1 <- readRDS( here::here( "data/rodeo/LTDB-2000.rds" ) )
d2 <- readRDS( here::here( "data/rodeo/LTDB-2010.rds" ) )
md <- readRDS( here::here( "data/rodeo/LTDB-META-DATA.rds" ) )

d1 <- dplyr::select( d1, - year )
d2 <- dplyr::select( d2, - year )

d <- merge( d1, d2, by="tractid" )
d <- merge( d, md, by="tractid" )

# keep only the urban tracts
d <- dplyr::filter( d, urban == "urban" )

# obtain NMTC data
NMTC_URL <- "https://raw.githubusercontent.com/DS4PS/cpp-528-spr-2020/master/labs/data/raw/NMTC/nmtc-sheet-01.csv"
nmtc <- read.csv( NMTC_URL, stringsAsFactors=F )

# obtain LIHTC data
LIHTC_URL <- "https://raw.githubusercontent.com/DS4PS/cpp-528-spr-2020/master/labs/data/raw/LIHTC/LIHTCPUB.csv"
lihtc <- read.csv( LIHTC_URL, stringsAsFactors=F )

# create a key that will allow us to obtain federal data for each tract ----
# remove anything not a number from the string
d$id2 <- gsub( "[^0-9]", "", d$tractid )

# fix IDs so they are match
d$id2 <- as.numeric( d$id2 )

# aggregate federal programs such that there is one record per tract ----
lihtc.dollars <-
  lihtc %>% 
  dplyr::filter( yr_alloc >= 2000 & yr_alloc <= 2010 ) %>%
  dplyr::group_by( fips2010 ) %>%
  dplyr::summarize( num.lihtc = dplyr::n(), lihtc.total = sum( allocamt, na.rm=T ) )

# need to convert from currency to numeric
# current format: 
# head( nmtc$QLICI.Amount )
# [1] "$300,000.00 "   "$1,008,750.00 " "$977,000.00 "

# remove dollar sign and commas
nmtc$amount <- gsub( "[,$]", "", nmtc$QLICI.Amount )

# head(  nmtc$amount  )
# "300000.00 "  "1008750.00 " "977000.00 "

# convert characters to numeric 
nmtc$amount <- as.numeric( nmtc$amount ) %>% round(0)

# head(  nmtc$amount  )
# [1]  300000 1008750  977000

nmtc.dollars <- 
  nmtc %>% 
  dplyr::filter( Origination.Year >= 2000 & Origination.Year <= 2010 ) %>%
  dplyr::group_by( X2010.Census.Tract ) %>% 
  dplyr::summarize( num.nmtc = dplyr::n(), nmtc.total = sum( amount, na.rm=T ) )

# merge federal data onto census tracts ----
d <- merge( d, nmtc.dollars, by.x="id2", by.y="X2010.Census.Tract", all.x=T )
d <- merge( d, lihtc.dollars, by.x="id2", by.y="fips2010", all.x=T )

# recode tracts that had no grants from NA to 0 ---
d$num.nmtc[ is.na(d$num.nmtc) ] <- 0
d$nmtc.total[ is.na(d$nmtc.total) ] <- 0

d$num.lihtc[ is.na(d$num.lihtc) ] <- 0 
d$lihtc.total[ is.na(d$lihtc.total) ] <- 0


# adjust 2000 home values for inflation 
mhv.00 <- d$mhmval00 * INFLATION_RATE  
mhv.10 <- d$mhmval12

# change in MHV in dollars
mhv.change <- mhv.10 - mhv.00


# drop low year 2000 median home values
# to avoid unrealistic growth rates.
#
# tracts with homes that cost less than
# $10,000 are outliers
# approximately 200 out of 59,000 cases 
# sum( mhv.00 < 10000 ) 
## [1] 200
mhv.00[ mhv.00 < 10000 ] <- NA

# change in MHV in percent
mhv.growth <- 100 * ( mhv.change / mhv.00 )

# store mini data frame to be used for descriptive statistics ----
df <- data.frame( MedianHomeValue2000=mhv.00, 
                  MedianHomeValue2010=mhv.10, 
                  MHV.Change.00.to.10=mhv.change,
                  MHV.Growth.00.to.12=mhv.growth )

# add variables to the main data frame ----
d$mhv.00 <- mhv.00
d$mhv.10 <- mhv.10
d$mhv.change <- mhv.change
d$mhv.growth <- mhv.growth 

# select a few variables ----
d <- dplyr::select( d, 
             
             tractid, cbsa, cbsaname,            # ids / units of analysis
             
             mhv.00, mhv.10, mhv.change, mhv.growth,    # home value 
             
             hinc00, hu00, own00, rent00,        # ses
             hinc12, hu10, own10, rent10,
             
             empclf00, clf00, unemp00, prof00,   # employment 
             empclf12, clf12, unemp12, prof12,
             
             dpov00, npov00,                     # poverty
             dpov12, npov12,
             
             ag25up00, hs00, col00,              # education 
             ag25up12, hs12, col12,
             
             pop00.x, nhwht00, nhblk00, hisp00, asian00,   # race
             pop10, nhwht10, nhblk10, hisp10, asian10,
             
             num.nmtc, nmtc.total,              # tax policy data
             num.lihtc, lihtc.total             # aggregated by census tract
             
) # end select


# create new variables ----
d <- 
  d %>%
  dplyr::mutate( 
    # 2000 variables
    p.white.00 = 100 * nhwht00 / pop00.x,
    p.black.00 = 100 * nhblk00 / pop00.x,
    p.hisp.00 = 100 * hisp00 / pop00.x, 
    p.asian.00 = 100 * asian00 / pop00.x,
    p.hs.edu.00 = 100 * (hs00+col00) / ag25up00,
    p.col.edu.00 = 100 * col00 / ag25up00,
    p.prof.00 = 100 * prof00 / empclf00,
    p.unemp.00 = 100 * unemp00 / clf00,
    pov.rate.00 = 100 * npov00 / dpov00,
    
    # 2010 variables
    p.white.10 = 100 * nhwht10 / pop10,
    p.black.10 = 100 * nhblk10 / pop10,
    p.hisp.10 = 100 * hisp10 / pop10, 
    p.asian.10 = 100 * asian10 / pop10,
    p.hs.edu.10 = 100 * (hs12+col12) / ag25up12,
    p.col.edu.10 = 100 * col12 / ag25up12,
    p.prof.10 = 100 * prof12 / empclf12,
    p.unemp.10 = 100 * unemp12 / clf12,
    pov.rate.10 = 100 * npov12 / dpov12 ) %>%
  # remove any NA or Inf values
  na.omit(use = "everything")

# inflation adjust income  ----
d$hinc00 <- INFLATION_RATE * d$hinc00

# create new variables by cbsa ----
d <-
  d %>%
  dplyr::group_by( cbsaname ) %>%
  dplyr::mutate( # metro rank of home value in 2000
    metro.mhv.pct.00 = dplyr::ntile( mhv.00, 100 ),
    # metro rank of home value in 2010
    metro.mhv.pct.10 = dplyr::ntile( mhv.10, 100 ),
    # median pay for metro area 2000
    metro.median.pay.00 = median( hinc00, na.rm=T ),
    # median pay for metro area 2010
    metro.median.pay.10 = median( hinc12, na.rm=T ),
    # tract rank in metro area for diversity (% non-white)
    metro.diversity.rank.00 = dplyr::ntile( (100-p.white.00), 100 ),
    # metro total population 2000
    metro.pop.total.00 = sum( pop00.x, na.rm=T ),
    # metro total population 2010
    metro.pop.total.10 = sum( pop10, na.rm=T ) ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate( # change in relative rank of mhv in metro area 2000-2010
    metro.mhv.pctile.change = metro.mhv.pct.10 - metro.mhv.pct.00,
    # growth in ave pay in metro
    metro.pay.change = metro.median.pay.10 - metro.median.pay.00,
    # metro population growth 2000-2010
    metro.pop.growth = ( metro.pop.total.10 - metro.pop.total.00 ) / metro.pop.total.00,
    # increase in the proportion of whites in tract 
    increase.p.white = p.white.10 - p.white.00  )

# Create a true/false code for recipient tracts ----
d$LIHTC <- ifelse( d$num.lihtc > 0, "YES", "NO" )
d$NMTC <- ifelse( d$num.nmtc > 0, "YES", "NO" )

# create a growth column within the data frame ----
d$growth <- d$mhv.growth
d$growth[ d$growth > 200 ] <- NA


d <- select(d, mhv.00, mhv.10, p.col.edu.00, p.col.edu.10, p.unemp.00, p.unemp.10, p.black.10, p.black.00, num.nmtc, nmtc.total,num.lihtc, lihtc.total)

d <- do.call(data.frame, lapply(d, function(x) {
  replace(x, is.infinite(x) | is.na(x), 0)
  }))


```



```{r message=FALSE, echo=FALSE}
#Save data to rodeo folder
saveRDS(d, here("data/rodeo/rodeo-date-mr.rds"))

```



```{r message=FALSE, echo=FALSE, include=FALSE}
#Load data set
readRDS(file = here("data/rodeo/rodeo-date-mr.rds"))
```




### **Difference-in-Difference Model**

#### The difference-in-difference estimate seems like the most robust option since it allows for pre-treatment differences as long as the parallel lines assumption is met. That is to say, the rates of growth for the treated and untreated tracts were similar in the period 1990 to 2000. Which is another way of saying we believe that C2-C1 is a good approximation of what the home value change would have been without the program.



### NMTC Analysis



```{r message=FALSE, echo=FALSE, results='asis'}

# store DV in both time periods
y1 <- log( d$mhv.00 )
y2 <- log( d$mhv.10 )

# store IV in both time periods
p.col00 <- as.numeric(log(d$p.col.edu.00))
p.col10 <- as.numeric(log(d$p.col.edu.10))
p.unemp00 <- as.numeric(log( d$p.unemp.00))
p.unemp10 <- as.numeric(log( d$p.unemp.10))
p.black00 <- as.numeric(log( d$p.black.00))
p.black10 <- as.numeric(log( d$p.black.10))

treat <- as.numeric( d$num.nmtc > 0 )

# add DV and IVs into one data frame per time period
d1 <- data.frame( y=y1, treat=treat, post=0, p.col=p.col00, p.unemp = p.unemp00, p.black = p.black00)
d2 <- data.frame( y=y2, treat=treat, post=1, p.col=p.col10, p.unemp = p.unemp10, p.black = p.black10)

# stack the time periods together into one data frame
d3 <- rbind( d1, d2 )
data_new <- d3                                   # Duplicate data
data_new[is.na(data_new) | data_new == "Inf" | data_new == "-Inf"] <- NA  # Replace NaN & Inf & -Inf with NA

###labels for knitting
PercentCollegeGrad <- data_new$p.col
PercentUnemployed <- data_new$p.unemp
PercentBlackPopulation <- data_new$p.black

m <- lm( y ~ p.col +p.unemp +p.black + treat + post + treat*post, data=data_new)

# view regression results
pander(summary( pander(summary(lm( y ~ PercentCollegeGrad +PercentUnemployed + PercentBlackPopulation + treat + post + treat*post, data=data_new))) ))

# display model results
stargazer::stargazer(m,
                     title  = "NMTC Treatment",
                     covariate.labels = c("Percent College Grads", "Percent Unemployed", "Percent Black Population"),
                     dep.var.labels = "Log Median Home Value",
                     type = S_TYPE,
                     digits = 2)
```



### LIHTC Analysis



```{r message=FALSE, echo=FALSE, results='asis'}

treat2 <- as.numeric( d$num.lihtc > 0 )

# add DV and IVs into one data frame per time period
d1 <- data.frame( y=y1, treat=treat2, post=0, p.col=p.col00, p.unemp = p.unemp00, p.black = p.black00)
d2 <- data.frame( y=y2, treat=treat2, post=1, p.col=p.col10, p.unemp = p.unemp10, p.black = p.black10)

# stack the time periods together into one data frame
d3 <- rbind( d1, d2 )
data_new <- d3                                   # Duplicate data
data_new[is.na(data_new) | data_new == "Inf" | data_new == "-Inf"] <- NA  # Replace NaN & Inf with NA



m <- lm( y ~ p.col +p.unemp +p.black + treat + post + treat*post, data=data_new)

###labels for knitting
PercentCollegeGrad <- data_new$p.col
PercentUnemployed <- data_new$p.unemp
PercentBlackPopulation <- data_new$p.black

# view regression results
pander(summary(lm( y ~ PercentCollegeGrad +PercentUnemployed + PercentBlackPopulation + treat + post + treat*post, data=data_new)))

# display model results
stargazer::stargazer(m,
                     title ="LIHTC Treatment",
                     covariate.labels = c("Percent College Grads", "Percent Unemployed", "Percent Black Population"),
                     dep.var.labels = "Log Median Home Value",
                     type = S_TYPE,
                     digits = 2)
```



### Parallel Lines Test



```{r message=FALSE, echo=FALSE, results='asis'}

# create a variable that identifies if a tract received NMTC funding
treat <- as.numeric( d$num.nmtc > 0 )

# store the year 2000 data
d1 <- data.frame( y=y1, treat=treat, post=0 )
# store the year 2010 data
d2 <- data.frame( y=y2, treat=treat, post=1 )

# stack the two time periods together
d3 <- rbind( d1, d2 )

# create a variable that identifies if a tract received NMTC funding
treat1 <- as.numeric( d$num.lihtc > 0 )

# store the year 2000 data
d4 <- data.frame( y=y1, treat=treat1, post=0 )
# store the year 2010 data
d5 <- data.frame( y=y2, treat=treat1, post=1 )

# stack the two time periods together
d6 <- rbind( d1, d2 )
d7 <- rbind( d4, d5 )

m <- lm( y ~ treat + post + treat*post, data=d6)
m2 <- lm( y ~ treat + post + treat*post, data=d7)

```



#### NMTC Regression Summary 
```{r, echo=FALSE}
# view regression results
pander(summary( m ))
```



#### LIHTC Regression Summary
```{r, echo=FALSE}
pander(summary(m2))
```


```{r, results='asis'}
# display model results
stargazer::stargazer(m,m2,
                     title  = "NMTC and LITHC Comparison",
                     dep.var.labels = "Log Median Home Value",
                     column.labels = c("NMTC", "LIHTC"),
                     type = S_TYPE,
                     digits = 2)
```


### Home value test


```{r message=FALSE, echo=FALSE}
home1 <- 200000
home2 <- 100000

rate.of.return <- .1/(1:10)

roi.1 <- home1*rate.of.return
roi.2 <- home2*rate.of.return 

plot( 1:10, roi.1, type="b", pch=19, bty="n",
      xlab="Time", ylab="Home Value", ylim=c(0,25000) )
points( 1:10, roi.2, type="b", pch=19 )

text( 5, roi.1[5]+100, "Home 1", pos=3, cex=1.5 )
text( 5, roi.2[5]+100, "Home 2", pos=3, cex=1.5 )

log.roi.1 <- log( roi.1 )
log.roi.2 <- log( roi.2 )

plot( 1:10, log.roi.1, type="b", pch=19, bty="n",
      xlab="Time", ylab="Home Value", main = "Logged ROI", ylim=c(0,15) )
points( 1:10, log.roi.2, type="b", pch=19 )

text( 5, log.roi.1[5], "Home 1", pos=3, cex=1.5 )
text( 5, log.roi.2[5], "Home 2", pos=3, cex=1.5 )

options(warn=-0)
```

```{r, echo=FALSE}
data.frame( time=1:10, value=roi.1, log.value=log.roi.1 ) %>% pander()
```

### True Growth Rate

```{r, echo=FALSE}
y1 <- roi.1[1]
y2 <- roi.1[2]


( y2 - y1 ) / y1
```

### Growth Rate Approximation

```{r, echo=FALSE}
log.y1 <- log.roi.1[1]
log.y2 <- log.roi.1[2]


log.y2 - log.y1
```


```{r, results='asis', echo=FALSE}
m <- lm( y ~ treat + post + treat*post, data=d3 )

# display model results
stargazer(m,
          title = "NMTC- Diff-In-Diff",
          type = S_TYPE,
          digits = 2)
```


#### Diff-in-diff model - equation

```{r, echo=FALSE}
# diff in diff model terms
b0 <- m$coefficients[1] 
b1 <- m$coefficients[2]
b2 <- m$coefficients[3]
b3 <- m$coefficients[4]
```





### C1 = B0 

```{r, echo=FALSE}
C1 <- b0
pander(exp( C1 ))
```

### C2 = B0 + B2 

```{r, echo=FALSE}
C2 <- b0 + b2
pander(exp( C2 ))
```

### T1 = B0 + B1 

```{r, echo=FALSE}
T1 <- b0 + b1
pander(exp( T1 ))
```

### T2 = B0 + B1 + B2 + B3 

```{r, echo=FALSE}
T2 <- b0+b1+b2+b3
pander(exp( T2 ))
```

### Counterfactual: C2-C1

```{r, echo=FALSE}
CF<- 196984 - 156581

pander(CF)
```

### Our Treatment Group: T2-T1

```{r, echo=FALSE}

TR <- 166560 - 120265
pander((TR))

```

### Treatment difference in difference (T2-T1) - (C2-C1)

```{r, echo=FALSE}
pander(TR-CF)
```


### Reflection

#### Looking at our regression models, it is obvious that some demographic variables impact our policy variables. College graduates have a significantly positive relationship with growth in median home value, while percent black population has significantly negative relationship with growth in median home value for both programs. As we know from other labs and research, the positive coefficient with percent unemployed is probably due to multicollinearity/interference from instrumental variables or missing variables. This is probably also true for the negative relationship with percent black population and median home value, as we know that black people do not cause home prices to fall. 

#### Because we are depending on a diff-in-diff model, the comparison of C and T values shows that our treatment group had a higher ROI than the control group. We are asserting that the counterfactual is represented by the ROI of our control group, or the growth of 'control' ROI from time 1 to time 2. We do not have a true counterfactual, but we can assert confidence that our counterfactual, by comparison, can isolate secular trends with minimal variance from unknown factors. We are extrapolating our counterfactual is comprised of: the post growth coefficient (secular trend, C2), the intercept of our untreated group (C1), the intercept of our treated group (T1) and the median home values post treatment within that group (T2)


#### We know that the base rate of increase in MHV was around 23%, and we can say the null hypothesis, that there was no difference in median home growth in MHV from 2000-2010 for our treatment group from the secular trend of 23%, can be rejected with confidence. 


#### Comparing regression coefficients, looking at graphed data, summary statistics and our diff-in-diff model, it is clear that NMTC performs above secular trends and above LIHTC. 


#### We test the parallel lines assumption by looking at the post regression coefficient in our regression table for both programs. We can see that the coefficient for post, meaning how much the median home value increased over the two stacked time periods, 1990-2000, and 2000-2010, are the same for both our treated and untreated (NMTC and LIHTC) groups. Because our y values, or dependent variables, are logged median home values, we can say that this represents median home value growth, and that the rate of growth is the same. We are effectively comparing apples to apples. This is butressed by our home values test, which visually illustrates this concept. After logging data, the two slopes are parallel, with different intercepts. 



#### In conclusion, it can be said that NMTC's model for low-income housing resulted in higher MHV growth than the secular trend, and it may be inferred that this leads to value improvement of neighborhoods, because home prices tend to reflect the quality of life value of the neighborhoods. However, because the NMTC approaach tended to have a 'middle of the road' approach to percent poverty tracts chosen, as seen by our data visualizations ("2000 Poverty Rate Comparison of \nRecipient and Non-Recipient Communities"), it is possible that they chose tracts strategically in areas that would have seen growth anyway, and less people in poverty actually benefited from these programs. Testing this hypothesis would require a different framework that was people-based, not place-based. 